---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r}
#Header code, setup required libraries
library(ggplot2)
library(plyr)
```

## Loading and preprocessing the data
```{r}
#read in
activity<-read.csv("activity.csv",colClasses=c("integer","Date","integer"))
#q1
stepsperday<-ddply(activity, c("date"),summarise,
                   totalsteps=sum(steps,na.rm=TRUE)
                   )
#q2
stepsper5min<-ddply(activity, c("interval"),summarise,
                    meansteps = mean(steps,na.rm=TRUE)
                    )

```

## What is mean total number of steps taken per day?

The mean total number of steps taken per day is `r mean(stepsperday$totalsteps, na.rm=TRUE)`.  The median number of steps taken per day is `r median(stepsperday$totalsteps)`(NA's omitted).

```{r stepshist}
stepshist<-ggplot(stepsperday,aes(x=totalsteps))+geom_histogram()+
  xlab("Total number of steps")+
  ggtitle("Histogram of total steps in one day")+
  theme_bw()
print(stepshist)
```

## What is the average daily activity pattern?
```{r daypattern,warning=FALSE}
dayline<-ggplot(stepsper5min,aes(x=interval,y=meansteps))+geom_line()+
  ggtitle("Average steps for each 5-min interval")+
  ylab("Mean steps")+
  theme_bw()
print(dayline)
```

Alternative daily activity pattern visualisation, raw points with a loess curve

```{r altdaypattern}
dayraw<-ggplot(activity,aes(x=interval,y=steps))+geom_point(alpha=.1)+geom_smooth()+
  ggtitle("Steps in each 5-min interval, raw points + loess curve")+
  theme_bw()
print(dayraw)
```

The five minute interval with the highest mean step-count is interval #`r stepsper5min[which(stepsper5min$meansteps==max(stepsper5min$meansteps)), "interval"]` with a mean of `r stepsper5min[which(stepsper5min$meansteps==max(stepsper5min$meansteps)), "meansteps"]` steps.  

## Imputing missing values

There are `r nrow(activity)-sum(complete.cases(activity))` incomplete records, unevenly distributed through the data.

```{r histincomplete}
hist(which(complete.cases(activity)),
     main="Count of complete cases (chronological order)",
     xlab="Observation number",
     ylab="Count of complete cases"
     )

```
 
Interpolation is done by using the average of the previous valid observation and the next valid observation, or the average for the relevant 5-min interval if there is no valid previous/next observation. This produces smooth activity-over-the-day lines for each individual day, but is not very fast.
 
```{r interpolation strategy}
#q3
step_interpolation <- function(rownumber){
  prevrow=rownumber;
  nextrow=rownumber;
  while(is.na(activity$steps[prevrow])){
    prevrow=prevrow-1
    if(prevrow<1)return(mean(activity[activity$interval==activity$interval[rownumber],"steps"],na.rm=TRUE))
  }
  while(is.na(activity$steps[nextrow])){
    nextrow=nextrow+1
    if(nextrow>nrow(activity))return(mean(activity[activity$interval==activity$interval[rownumber],"steps"],na.rm=TRUE))
  }
  return(
    (activity$steps[prevrow]+activity$steps[nextrow])/2
  )
}

activity_guessNA <-activity
for(n in 1:nrow(activity)){
  if(is.na(activity$steps[n])){
    activity_guessNA$steps[n]=step_interpolation(n);
  }
}
```

I know, this is a density plot not a histogram, but the meaning is the same and I didn't want to superimpose two histograms. The imputed dataset has (relatively) fewer zeros, the original data is peppered with lone zeros and the imputation strategy above just doesn't reproduce this pattern. Most of the imputed entries appear to have been added in the most commonly occuring range.

```{r guesscompare}
stepsperday2<-merge(
  ddply(activity_guessNA, c("date"),summarise,
        guesstotalsteps=sum(steps,na.rm=TRUE)
  ),
  stepsperday,
  by="date"
)

guesscheck<-ggplot(stepsperday2,aes(x=totalsteps))+
  geom_density()+
  geom_density(aes(x=guesstotalsteps,color="Imputed"))+
  ggtitle("Density plot comparing raw and NA-imputed activity datasets")+
  xlab("total steps")+
  theme_bw()
print(guesscheck)
```

Here's the histogram for my fellow pedants:
```{r imputedhist}
forpeoplewhoreallywanttoseeahistogram<-ggplot(stepsperday2,aes(x=guesstotalsteps))+
    geom_histogram()+
  ggtitle("Histogram of total number of steps per day after missing values imputed")+
    theme_bw()
print(forpeoplewhoreallywanttoseeahistogram)
```

The mean and median total steps are `r mean(stepsperday2$totalsteps,na.rm=TRUE)` and `r median(stepsperday2$totalsteps,na.rm=TRUE)`, for the NA-imputed data the mean and median are  `r mean(stepsperday2$guesstotalsteps,na.rm=TRUE)` and `r median(stepsperday2$guesstotalsteps,na.rm=TRUE)`. 

## Are there differences in activity patterns between weekdays and weekends?

Looks like activity is higher on the weekends, particularly in the middle of the day, although it is lower early in the morning just after waking.

```{r weekends}
paindays= c("Monday","Tuesday","Wednesday","Thursday","Friday")

activity_guessNA$weekday<-as.factor(ifelse(weekdays(activity_guessNA$date)%in%paindays,"weekday","weekend"))

stepsperinterval.weekdaysplit<-ddply(activity_guessNA, c("interval","weekday"),summarise,
                    meansteps = mean(steps,na.rm=TRUE)
)

weekdayplot<-ggplot(stepsperinterval.weekdaysplit,aes(x=interval,y=meansteps))+
  facet_wrap(~weekday,nrow=2,ncol=1)+
  geom_line()+
  theme_bw()+
  ggtitle("Mean steps over each 5min interval split by weekday/weekend")+
  ylab("Mean steps")+
  xlab("Interval number")
print(weekdayplot)
```